Transcriptional changes in the reproductive axis of parenting pigeons

### 2020-05-18 docs/manuscript/elife-v1.tex
now torking

\emph{Characterization of parental care}

To characterize transcriptional activity, we first examined broad
patterns of variation in gene expression in the hypothalamus, pituitary,
and gonads of male and female rock doves across a series of timepoints
that encapsulate important transitions in reproduction and and parental
care (Fig. \ref{figure:fig1}a). t-Distributed Stochastic Neighbor
Embedding (t-SNE) is a nonlinear, dimension reduction technique is used
to visualize high-dimensional data by giving each datapoint a location
in a two-dimensional map (van der Maaten and Hinton 2008). t-SNE is
widely using in single-cell data process for identify related cell
types. Here, we use t-SNE to identify variation associated with tissue,
sex, and treatment (Fig. \ref{figure:fig1}j).

Our results suggest that tissue is the greatest source of variation in
the dataset, with hypothalamic, pituitary, and gonadal samples all
forming distinct clusters, with only a few pituitary and gonad samples
falling outside the 95\% confidence interval (Fig. \ref{figure:fig1}b).
Sex-differences are most prounounced in the gonads with male and female
samples form two discrete clusters (Fig. \ref{figure:fig1}c). Treatment
accounts for much less of the variation than tissue and sex, and even
after removing sex as a variatble, treatment effects in all three
tissues are hard to visualize simultatenously (Fig.
\ref{figure:fig1}d,e), so we next looked at each tissue and sex
separately. In the hypothalamus, control samples form tight clusters
that stands out from all other timepoints that are largely overlapping
in space (Fig. \ref{figure:fig1}h). The pituitary shows the strongest
treatment effect with one cluster of the ``earlier'' parental timepoints
(nest-building, lay, incubation days 3 and 9), a second cluster of
``later'' parental timpoints (incubation day 17, hatch, and nestling
care days 5 and 9; Fig. \ref{figure:fig1}i). Like the hypothalamus, in
the gonads, the control timepoints appear to be much less variable that
the parental timepoints, but there is no clear separation of gonadal
samples by parental stage (Fig. \ref{figure:fig1}j). While the tSNE
analysis reveals broad patterns across all levels of the experiment,
additional tools are needed it hone in on the specific differences
associated with parental care in each tissue. Differential gene
expression analysis can shed insight into specific differences between
treatment groups and thus deepen our understand of the transitions
between reproduction and parental care behaviors. We used DESEq2 to
identify genes whose expression changed significantly between sequential
timepoints in each tissue and sex separately (Fig. \ref{figure:fig2}a).

In the hypothalamus only two sequential stage comparisons results in a
sizable change in expression; in females, 1927 genes changes their
expression betwen hatch to nesting care day 5 (with 1076 decreasing and
851 increased their expression), and in the male hypothalamus,87 genes
were diferentially expressed bewteen incubation days 9 and 17 (with 38
decreasing and 49 increasing their expression (Fig.
\ref{figure:fig2}b)). In the pituitary, differential gene expression
differenes are greatest when compariing incubation days 9 and 17, and
this difference is more pronounced in females (Fig. \ref{figure:fig2}c).
The next largest difference in the pituitary is between hatch and n5.
The female putitary shows moderate chagnes from nest-bilding to lay to
incuabation day 3 (Fig. \ref{figure:fig2}c). The male pituitary changes
more that the female pitutiary changes between nestling care days 5 and
9 (Fig. \ref{figure:fig2}c). About 2000 genes also change their
expression in the female pituitary and about 425 in the male pituitary
when comparing incubation days 9 and 17 (Fig. \ref{figure:fig2}). In the
female gonads, 270 genes are decreased while only 56 are increased when
comparing incubation days 9 and 17(Fig. 2). Many more significant
changes in gene expression in the female gonad occur between
nest-building and lay and between lay and clutch completion (incubation
day 3) than in any other comparison (Fig. \ref{figure:fig2}).

\emph{Internal physiology versus external stimuli hypothesis}

We show that samples from the pituitary on incubation day 17 cluster
more closely with hatch and nestling care day 5 samples than with
incubation days 3 and 9. This suggest that, in the pituitary, internal
mechanisms that prepare for chick arrival have the greatest effect on
gene expression rather than suggesting that gene expression responds to
changes in the external enviornament.

Blah \ref{figure:fig3}, blah, \ref{figure:fig4} blah.

\emph{Potential implications for cancer research}

These results confirmed prolactin (PRL) gene expression fluctuates
throughout parental care in a manner that is consistent with its role in
promoting lactation and maintaining parental behaviors. Interestingly,
we identified about 100 genes whose expression was correlated with PRL,
including (BRCA1, which is associated with breast cancer). Additionally,
we identified thousands of genes whose expression changed over the
course of parental care. Exploring the relationship between PRL and
other differentially expressed genes could provide important insights
into the \texttt{symphony} of gene expression that regulates behavior at
the organismal and cellular level.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig1-1}
  \caption{Broad patterns of variation in hypothalamic, pituitary, and gonadal gene expression in male and female rock doves across reproductive and parental transitions. a) Experimental design. b) t-SNE plot highlighting tissue differences in gene expression. Each dot represents 1 sample and is colored by the source tissue (hypothalamus: orange, pituitary: teal, gonads: purple). c) t-SNE plot highlighting sex differences in tissue and sex, as shown in light grey (females) and dark grey (males). t-SNE plots highlight treatment effects in females (d), males (e), the hypothalamus (h), pituitary (i), and gonads (j), with samples colored by parental stage (control: red or , nest-building: orange, lay: yellow; inc.d3: light green, inc.d9: green, inc.d17: dark green, hatch: dark blue, n5: blue, and n9: light blue.}
  \label{figure:fig1}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig3-1}
  \caption{Another figure}
  \label{figure:fig3}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/supplfig3}
  \caption{R shiny, image 1}
  \label{figure:supplfig3}
\end{figure}

>>>>

\emph{Data-driven characterization of reproductive and parental stage
groups.}

We first examined broad patterns of variation in gene expression in the
hypothalamus, pituitary, and gonads of male and female rock doves across
a series of timepoints that encapsulate important transitions in
reproduction and and parental care (Fig. 1A). t-Distributed Stochastic
Neighbor Embedding (t-SNE) is a nonlinear, dimension reduction technique
is used to visualize high-dimensional data (van der Maaten and Hinton
2008) and is widely used in in RNA sequencing to identify related and
distinct cell types (refs). Differential expespression profiling is
useful for pair-wise comparisons of differentially expressed genes
(DEGs) with positive log fold change (+LFC) in one group compared to
another. Combined, the provide stronger support for similarities and
differences between cell-types and treatment groups in multi-dimential
datasets. First, the data supports that the hypothalamus, pituitary, and
gonads have distinct gene expression profiles as evidenced by little
overlap in t-SNE space and an average of 13,000 DEGs between each tissue
(Fig. 1B). Second, the data supports that sex-differences are much more
promient in the gonads than in the brain (13,000 versus 2,000 DEGs)
(Fig. 1C). Because of these widespread tissue and sex-differences,
further ananlyses were conducted separately for each tissue and sex
(Fig. 1D, Suppl. Fig 1). In all the tissues, the control samples are
quiet different from all the other timepoints samples, and only a
fraction of the transcriptome is altered during the course of parental
care at the timepoints sampled. The most notable differences male and
female hpothalamic control samples are very different from on the rest
(\textasciitilde{}7000 DEGs, and only the two nestling care stages (n5
and n9)) show differential expression of more than 1,000 genes. In the
pituitary, late inubcation and hatch are quite simlilar to each other
and distinct from the earlier incubation stages stages, and temporal
changes in gene expression appears to be less variable in males than
females. In the female gonads there appear to be more changes in gene
expression early one, between nest building and early incuation, but the
male gonads show very little variation associated with parental stage.

\emph{Hypothesis-driven analysis of candidate parental-care genes.}

\begin{itemize}
\tightlist
\item
  Champagne Curley Genes (Fig 2)
\item
  GO genes (Fig 2 Suppl Fig 1)
\item
  Cancer genes (Fig 2 Suppl Fig 2)
\item
  Shiny genes (Fig 2 Suppl Fig 3)
\end{itemize}

@@

\textbackslash{}begin\{figure\}{[}h{]} \centering
\includegraphics[width=1.0\textwidth]{../../figures/fig1-1}
\textbackslash{}caption\{Broad patterns of variation in hypothalamic,
pituitary, and gonadal gene expression in male and female rock doves
across reproductive and parental transitions.** A) Experimental design.
B) The t-SNE plot shows all 636 samples as points in space colored by
tissue with ellipses drawn around show 95\% confidence interval, and the
bar plots show the total number of differentially expressioned genes
(DEGs) whose expression was significantly higher in one tissue versus
the other. C) This t-SNE plot shows the same data points colored by sex
while the ellipses delinate the source tissue; bar plots show the DEGs
between the sexes for each tissue. D) These t-SNE and bar plots provide
insight into the temporal differences in gene expression within each
tissue for each sex. Here, the bar plots highlight the number of DEGs
between each sequential time point. For comparisons to controls, see
fupplemental figure 1.\} \label{figure:fig1}
\textbackslash{}end\{figure\}

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/supplfig-1-1}
  \caption{Summary of the total number of differentially expressed genes (DEGs) from pair-wise comparison using DESeq2.** Bar plots showing the total number of differentially expressed genes (p < 0.1) and colored by the direction of increased expression (or positive log fold change (+LFC))  are shown for three different analysis: relative to the control group (A), the nest-building (bldg) group (B), and the previous sampling timepoint (C).}
  \label{figure:supplfig1}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig3-1}
  \caption{Another another figure}
  \label{figure:fig3}
\end{figure}

@@

### 2020-05-18 docs/manuscript/elife_v1.Rmd
now torking

_Characterization of parental care_

To characterize transcriptional activity, we first examined broad patterns of variation in gene expression in the hypothalamus, pituitary, and gonads of male and female rock doves across a series of timepoints that encapsulate important transitions in reproduction and and parental care (Fig. \ref{figure:fig1}a). t-Distributed Stochastic Neighbor Embedding (t-SNE) is a nonlinear, dimension reduction technique is used to visualize high-dimensional data by giving each datapoint a location in a two-dimensional map (van der Maaten and Hinton 2008). t-SNE is widely using in single-cell data process for identify related cell types. Here, we use t-SNE to identify variation associated with tissue, sex, and treatment (Fig. \ref{figure:fig1}j).

Our results suggest that tissue is the greatest source of variation in the dataset, with hypothalamic, pituitary, and gonadal samples all forming distinct clusters, with only a few pituitary and gonad samples falling outside the 95% confidence interval (Fig. \ref{figure:fig1}b). Sex-differences are most prounounced in the gonads with male and female samples form two discrete clusters (Fig. \ref{figure:fig1}c). Treatment accounts for much less of the variation than tissue and sex, and even after removing sex as a variatble, treatment effects in all three tissues are hard to visualize simultatenously (Fig. \ref{figure:fig1}d,e), so we next looked at each tissue and sex separately. In the hypothalamus, control samples form tight clusters that stands out from all other timepoints that are largely overlapping in space (Fig. \ref{figure:fig1}h). The pituitary shows the strongest treatment effect with one cluster of the “earlier” parental timepoints (nest-building, lay, incubation days 3 and 9), a second cluster of “later” parental timpoints (incubation day 17, hatch, and nestling care days 5 and 9; Fig. \ref{figure:fig1}i). Like the hypothalamus, in the gonads, the control timepoints appear to be much less variable that the parental timepoints, but there is no clear separation of gonadal samples by parental stage (Fig. \ref{figure:fig1}j). While the tSNE analysis reveals broad patterns across all levels of the experiment, additional tools are needed it hone in on the specific differences associated with parental care in each tissue. Differential gene expression analysis can shed insight into specific differences between treatment groups and thus deepen our understand of the transitions between reproduction and parental care behaviors. We used DESEq2 to identify genes whose expression changed significantly between sequential timepoints in each tissue and sex separately (Fig. \ref{figure:fig2}a).

In the hypothalamus only two sequential stage comparisons results in a sizable change in expression; in females, 1927 genes changes their expression betwen hatch to nesting care day 5 (with 1076 decreasing and 851 increased their expression), and in the male hypothalamus,87 genes were diferentially expressed bewteen incubation days 9 and 17 (with 38 decreasing and 49 increasing their expression (Fig. \ref{figure:fig2}b)). In the pituitary, differential gene expression differenes are greatest when compariing incubation days 9 and 17, and this difference is more pronounced in females (Fig. \ref{figure:fig2}c). The next largest difference in the pituitary is between hatch and n5. The female putitary shows moderate chagnes from nest-bilding to lay to incuabation day 3 (Fig. \ref{figure:fig2}c). The male pituitary changes more that the female pitutiary changes between nestling care days 5 and 9  (Fig. \ref{figure:fig2}c).   About 2000 genes also change their expression in the female pituitary and about 425 in the male pituitary when comparing incubation days 9 and 17 (Fig. \ref{figure:fig2}). In the female gonads, 270 genes are decreased while only 56 are increased when comparing incubation days 9 and 17(Fig. 2). Many more significant changes in gene expression in the female gonad occur between nest-building and lay and between lay and clutch completion (incubation day 3) than in any other comparison (Fig. \ref{figure:fig2}).

_Internal physiology versus external stimuli hypothesis_

We show that samples from the pituitary on 
                     incubation day 17 cluster more closely with hatch and nestling care 
                     day 5 samples than with incubation days 3 and 9. This suggest that, 
                     in the pituitary, internal mechanisms that prepare for chick arrival 
                     have the greatest effect on gene expression rather than suggesting 
                     that gene expression responds to changes in the external enviornament.

Blah, blah, \ref{figure:fig4} blah.

_Potential implications for cancer research_

These results confirmed prolactin (PRL) gene expression fluctuates throughout parental care 
in a manner that is consistent with its role in promoting lactation 
                and maintaining parental behaviors.
                Interestingly, we identified about 100 genes whose expression 
                was correlated with PRL,
                including (BRCA1, which is associated with breast cancer).
                Additionally, we identified thousands of genes whose expression 
                changed over 
                the course of parental care.
                Exploring the relationship between PRL and other differentially 
                expressed genes could provide important insights into the `symphony` of 
                gene expression 
                that regulates behavior at the organismal and cellular level.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig1-1}
  \caption{Broad patterns of variation in hypothalamic, pituitary, and gonadal gene expression in male and female rock doves across reproductive and parental transitions. a) Experimental design. b) t-SNE plot highlighting tissue differences in gene expression. Each dot represents 1 sample and is colored by the source tissue (hypothalamus: orange, pituitary: teal, gonads: purple). c) t-SNE plot highlighting sex differences in tissue and sex, as shown in light grey (females) and dark grey (males). t-SNE plots highlight treatment effects in females (d), males (e), the hypothalamus (h), pituitary (i), and gonads (j), with samples colored by parental stage (control: red or , nest-building: orange, lay: yellow; inc.d3: light green, inc.d9: green, inc.d17: dark green, hatch: dark blue, n5: blue, and n9: light blue.}
  \label{figure:fig1}
\end{figure}

@@

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig5-1}
  \caption{Hypothesis-driven data exploration and software development.}
  \label{figure:fig5}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/supplfig1}
  \caption{R shiny, image 1}
  \label{figure:supplfig3}
\end{figure}

>>>>

_Data-driven characterization of reproductive and parental stage groups._

We first examined broad patterns of variation in gene expression in the hypothalamus, pituitary, and gonads of male and female rock doves across a series of timepoints that encapsulate important transitions in reproduction and and parental care (Fig. 1A). t-Distributed Stochastic Neighbor Embedding (t-SNE) is a nonlinear, dimension reduction technique is used to visualize high-dimensional data (van der Maaten and Hinton 2008) and is widely used in in RNA sequencing to identify related and distinct cell types (refs). Differential expespression profiling is useful for pair-wise comparisons of differentially expressed genes (DEGs) with positive log fold change (+LFC) in one group compared to another. Combined, the provide stronger support for similarities and differences between cell-types and treatment groups in multi-dimential datasets. First, the data supports that the hypothalamus, pituitary, and gonads have distinct gene expression profiles as evidenced by little overlap in t-SNE space and an average of 13,000 DEGs between each tissue (Fig. 1B). Second, the data supports that sex-differences are much more promient in the gonads than in the brain (13,000 versus 2,000 DEGs) (Fig. 1C). 
Because of these widespread tissue and sex-differences, further ananlyses were conducted separately for each tissue and sex (Fig. 1D, Suppl. Fig 1). In all the tissues, the control samples are quiet different from all the other timepoints samples, and only a fraction of the transcriptome is altered during the course of parental care at the timepoints sampled. The most notable differences male and female hpothalamic control samples are very different from on the rest (~7000 DEGs, and only the two nestling care stages (n5 and n9)) show differential expression of more than 1,000 genes. In the pituitary, late inubcation and hatch are quite simlilar to each other and distinct from the earlier incubation stages stages, and temporal changes in gene expression appears to be less variable in males than females. In the female gonads there appear to be more changes in gene expression early one, between nest building and early incuation, but the male gonads show very little variation associated with parental stage.

_Hypothesis-driven analysis of candidate parental-care genes._

- Champagne Curley Genes (Fig 2)
- GO genes (Fig 2 Suppl Fig 1)
- Cancer genes (Fig 2 Suppl Fig 2)
- Shiny genes (Fig 2 Suppl Fig 3)

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig1-1}
  \caption{Broad patterns of variation in hypothalamic, pituitary, and gonadal gene expression in male and female rock doves across reproductive and parental transitions.** A) Experimental design. B) The t-SNE plot shows all 636 samples as points in space colored by tissue with ellipses drawn around show 95% confidence interval, and the bar plots show the total number of differentially expressioned genes (DEGs) whose expression was significantly higher in one tissue versus the other. C) This t-SNE plot shows the same data points colored by sex while the ellipses delinate the source tissue; bar plots show the DEGs between the sexes for each tissue. D) These t-SNE and bar plots provide insight into the temporal differences in gene expression within each tissue for each sex. Here, the bar plots highlight the number of DEGs between each sequential time point. For comparisons to controls, see fupplemental figure 1.}
  \label{figure:fig1}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/supplfig-1-1}
  \caption{Summary of the total number of differentially expressed genes (DEGs) from pair-wise comparison using DESeq2.** Bar plots showing the total number of differentially expressed genes (p < 0.1) and colored by the direction of increased expression (or positive log fold change (+LFC))  are shown for three different analysis: relative to the control group (A), the nest-building (bldg) group (B), and the previous sampling timepoint (C).}
  \label{figure:supplfig1}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=1.0\textwidth]{../../figures/fig3-1}
  \caption{Another another figure}
  \label{figure:fig3}
\end{figure}

@@
